---
title: "Week 4 Pre-Lecture Instructions"
author: "Ronnie Bailey-Steinitz"
date: "2025-10-16"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

# Preparing for Week 4

Before we start working with the Week 4 materials, make sure your folder and project are correctly set up.  
This ensures that all file paths work with the `here()` package and that your code runs smoothly when knitted.


## 1. Create a Week 4 Folder

Inside your main project directory - **`R-ecology-workshop`** - create a new folder called **`Week 4`**.  

Your directory should now look like this:

R-ecology-workshop/  
  |  
  ___ Week 1/  
  |  
  ___ Week 2/  
  |  
  ___ Week 3/  
  |  
  ___ Week 4/ *<- move downloaded files into here*  

> *Important:* Move your downloaded files **directly into this Week 4 folder** -  
> do **not** leave them inside a nested “Downloads” or “Week 4” subfolder.

  
## 2. Download the Week 4 Materials

All Week 4 files are available in the shared \href{https://drive.google.com/drive/folders/...}{\textcolor{blue}{*Google Drive Week 4 folder*}}.


Download and save the following **five files** into your new **`R-ecology-workshop/Week 4/`** folder:

- `climate.csv`  
- `forest_area.csv`  
- `country_area.csv`  
- `countries_continents.csv`  
- `Week 4 Joining and Reshaping BLANK.Rmd` *(blank file to get you started)*


## 3. Open Your Project and File

1. In RStudio, open your main project: **`R-ecology-workshop.Rproj`**
2. In the *Files* pane, navigate to your new **Week 4** folder.  
3. Open the file **`Week 4 Joining and Reshaping BLANK.Rmd`** - this is where we’ll work today.


## 4. Load the Week 4 Datasets

Once each dataset has been placed inside your Week 4 folder, load them with:
```{r, warning=FALSE, eval=FALSE}

library(tidyverse)
library(here)
library(janitor)

climate <- read_csv(here("Week 4/climate.csv")) %>% janitor::clean_names()
forest <- read_csv(here("Week 4/forest_area.csv")) %>% janitor::clean_names()
area <- read_csv(here("Week 4/country_area.csv")) %>% janitor::clean_names()
cont <- read_csv(here("Week 4/countries_continents.csv")) %>% janitor::clean_names()

```

## About the datasets
- `climate.csv` - Annual mean temperature per country (long format).
- `forest_area.csv` - Forested area per country by year (wide format; columns are years).
- `country_area.csv` - Static total land area (km²) per country (two columns).
- `countries_continents.csv` - Lookup table linking each country to its continent.

Once everything is loaded, we are ready to start joining the datasets together!


## Important! **Knitting to PDF**
If you've had issues with knitting to PDF, try running the following two commands in your console: 
```{r, eval=FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```
... one line at a time. The second line (`install_tinytex()`) will take some time ( approx. 30-60 minutes! depending on the internet connection), so be patient and let it finish installing.
