---
title: "Week 1: Is My Data Clean? Exploring, Diagnosing, and Visualizing Problems"
author: "Ronnie Steinitz"
output: html_document
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 200
---

We will be working using R Markdown files to save our code or script. An R Markdown file (saved as **'filename.Rmd'**) is a special type of document that lets you combine your R code, results (like tables and plots), and your written explanations all in one place.  

It’s especially useful because it creates a clear, reproducible record of your analysis that you (or others) can easily follow and re-run later. Unlike plain code files, R Markdown helps you communicate your
work like a report, not just as raw code.  

An Rmd consists of intermingled information narratives and code.  


# Skills Learning -- Lecture
## 0. Rmarkdown structure - code chunk

```{r}

x <- 5  # radius of a circle

```

> Useful commands:
- To "comment out" (make text grey and not run the code): Ctrl + Shift + C  
- To create a new code chunk: Ctrl + Alt + i  

Alright, let's get started.   

## Load Required Packages

```{r, message=FALSE, warning=FALSE}

# install.package("tidyverse")
# install.package("janitor")

# Load tidyverse for data wrangling and visualization


```


## 1. Load and Preview the Dataset

```{r}

# check that you are in the correct working directory, the R-Ecology_Workshop folder
# If not, set your working directory to the correct folder (e.g., R-Ecology-Workshop), where all of your scripts and data will be saved. For example: 
# setwd("/Users/rsteinitz/Documents/github/R Data Analysis Course")


# Load dataset



```


## 2. Diagnosing Data Types and Structure
> Remember, to create a new code chunk: Ctrl + Alt + i  








------------------------------------------------------------------------

# Skills Application -- Lab 
## 6. Practice Prompts 

Each student should now begin a **new R Markdown file** in the same project and try the following with either their own dataset or a provided one:

If you are working on **your own dataset**, then create a Markdown file which you will be using continuously and adding to each week. 
- Name it: *Lastname_Firstname_Data*  

If you are working on an **example dataset**, then create a Markdown file just for this week.
- Name it: *Lastname_Firstname_Week1*  

Then, begin the lab activity:
0. Import the dataset into the environment and save as an object. 

1.  Identify which variables have missing values.  

2.  Check for misclassified columns (e.g., text stored as factors, numbers stores as text).

3.  Convert at least one column to the correct type.  

4.  Get summary statistics for two variables of different classes.  

5.  Create one bar plot (categorical variable) and one histogram (numeric variable).  

6.  Add a title and axis labels to each plot.  

7.  Knit the script into an HTML file, which will be saved to your working directory



------------------------------------------------------------------------

## Summary

This week we:  
- Loaded and explored a new dataset   
- Investigated data types and missing values   
- Visualized numeric and categorical variables  
- knitted code into coding documentation  

Next week, we’ll start wrangling our data using `filter()`, `select()`, and `mutate()` from `dplyr`.   


# Optional - additional functions
To explore on your own time, ahead of next week
```{r}

# See which individual records have the most missing data.
# Number of missing values per row
data %>%
  mutate(row_missing = rowSums(is.na(.))) %>%
  arrange(desc(row_missing)) %>%
  head()

# Find out whether certain groups (e.g., sexes or species) are more likely to have missing values.
# Compare missing values in body mass across species
data %>%
  group_by(species) %>%
  summarise(missing_body_mass = sum(is.na(body_mass_g)))

# Customize colors in a bar plot using scale_fill_manual().
ggplot(data, aes(x = species, fill = species)) +
  geom_bar() +
  scale_fill_manual(values = c("darkgreen", "darkorange", "steelblue")) +
  labs(title = "Species Counts (Colored)", x = "Species", y = "Count")

# Use facet_wrap() to compare distributions across groups.
ggplot(data, aes(x = flipper_length_mm)) +
  geom_histogram(binwidth = 2, fill = "lightblue", color = "black") +
  facet_wrap(~ species) +
  labs(title = "Flipper Length by Species", x = "Flipper Length (mm)", y = "Count")

# Look at how many penguins exist per island and species.
# desc() dictates the *descending* order
data %>%
  count(island, species) %>%
  arrange(desc(n))

```

------------------------------------------------------------------------

![Palmer
Penguins!](https://education.rstudio.com/blog/2020/07/palmerpenguins-cran/gorman-penguins.jpg)

![Palmer
Map](https://education.rstudio.com/blog/2020/07/palmerpenguins-cran/antarctica-map.png)


